\texttt{ensmallen}: a flexible C++ library for efficient function optimization

### 2018-10-15 doc/mlsys_paper/paper.tex
clarifications, sharpening and notes for further improvements

To demonstrate the benefits of these optimizations that {\tt ensmallen} can take
advantage of, we compare {\tt ensmallen}'s performance to a number of other
optimization frameworks, including some that use automatic differentiation.

First, we consider the added overhead of other frameworks.  For our experiment,
we choose the simple and popular Rosenbrock function~\cite{Rosenbrock1960}:
$f([x_1, x_2]) = 100 (x_2 - x_1^2)^2 + (1 - x_1^2)$.  For the optimizer, we use
simulated annealing~\cite{kirkpatrick1983optimization},
a gradient-free optimizer.  Simulated annealing will call the objective function
very many times; for each simulation we limit the optimizer to 100k
objective evaluations.  We compare four frameworks for this task:

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\tt ensmallen}
  \item {\tt scipy.optimize.anneal}, from scipy 0.14.1~\cite{jones2014scipy}
  \item {\tt Optim.jl}'s simulated annealing implementation with Julia
1.0.1~\cite{mogensen2018optim}
  \item {\tt samin} from GNU Octave's {\tt optim} package~\cite{octave}
\footnote{Another option here might be {\tt simulannealbnd()} from MATLAB's
Global Optimization Toolkit.  However, no license was available for these
simulations.}
\end{itemize}
\vspace*{-0.3em}

Table~\ref{tab:rosenbrock_results} shows the average runtime
across 10 trials for $100000$ iterations of simulated annealing with each
implementation.  The simulations were run on a % TODO: what are Marcus's
% system specs?
{\tt \$COMPUTER}.
Of these frameworks, only Julia and {\tt ensmallen} are able to avoid the
function pointer dereference and take advantage of inlining and related
optimizations---and this difference is strongly reflected in the results.

Next, we consider the same machine learning example we have been working with
through the paper: linear regression.  For this task we will use the first-order
L-BFGS optimizer~\cite{zhu1997algorithm},
so for {\tt ensmallen} we will implement two versions: one with {\tt Evaluate()}
and {\tt Gradient()}, and one with only {\tt EvaluateWithGradient()}.  The code
for each of these is the same as shown earlier.  In addition, we have more
options for Julia: we can implement only the objective function and allow either
Julia's {\tt
Calculus.jl}\footnote{\url{https://github.com/JuliaMath/Calculus.jl}} or {\tt
ForwardDiff.jl}~\cite{RevelsLubinPapamarkou2016} package to automatically
compute the gradient.  Therefore, we compare against the following:

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\tt ensmallen}: only {\tt EvaluateWithGradient()} implemented
  \item {\tt ensmallen}-2: both {\tt Evaluate()} and {\tt Gradient()} separately
implemented
  \item {\tt Calculus.jl}: uses the default {\tt Calculus.jl} for
computing the gradient with {\tt Optim.jl}
  \item {\tt ForwardDiff.jl}: uses {\tt ForwardDiff.jl} for computing
the gradient with {\tt Optim.jl}
  \item {\tt Optim.jl}: implements the gradient manually
  \item {\tt scipy}: implements both objective and
gradient
  \item {\tt bfgsmin}: uses GNU Octave's {\tt optim} package
\end{itemize}
\vspace*{-0.3em}

Results for different data sizes are shown in Table~\ref{tab:lbfgs}.  For each
implementation, L-BFGS was allowed to run for only $10$ iterations and never
converged in fewer iterations.  The datasets trained on were highly noisy random
data with a slight linear pattern---but note that the exact data is not relevant
for our experiments here, only its size.  Runtimes are again reported as the
average across 10 runs.

\begin{table}
\begin{center}
\begin{tabular}{lccccc}
\toprule
{\em algorithm} & $d$: 100, $n$: 1k & $d$: 100, $n$: 10k & $d$: 100, $n$:
100k & $d$: 1k, $n$: 100k \\
\midrule
% TODO: this was only one trial on Ryan's desktop!
{\tt ensmallen} & {\bf 0.001s} & {\bf 0.006s} & {\bf 0.139s} & {\bf 1.698s} \\
{\tt ensmallen}-2 & 0.002s & 0.008s & 0.204s & 2.497s \\
{\tt Calculus.jl} & 0.172s & 0.960s & 27.428s & 2535.507s \\
{\tt ForwardDiff.jl} & 0.553s & 1.162s & 6.665s & 778.250s \\
{\tt Optim.jl} & 0.007s & 0.025s & 0.390s & 4.107s \\
{\tt scipy.optimize} & 0.002s & 0.010s & 0.198s & 2.022s \\
{\tt bfgsmin} & 0.215s & 1.705s & 15.998s & ??? \\
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for linear regression function with several different dataset
sizes.  All Julia runs do not count compilation time.}
\label{tab:lbfgs}
\end{table}

We can see that the use of {\tt EvaluateWithGradient()} yields considerable
speedup of almost 1.5x over the {\tt ensmallen} implementation with both the
objective and gradient implemented separately.  In addition, although the
automatic differentiation support makes it easier for the user to write their
code, it is very clear that the code generated by the automatic differentiator
is far from optimal, with runs that take up to three orders of magnitude longer!

>>>>

To demonstrate the benefits of the metaprogramming based code optimizations
that {\tt ensmallen} can exploit,
we compare the performance of {\tt ensmallen} with several other
optimization frameworks, including some that use automatic differentiation.

First, we consider the added overhead of other frameworks.  For our experiment,
we use the simple and popular Rosenbrock function~\cite{Rosenbrock1960}:
$f([x_1, x_2]) = 100 (x_2 - x_1^2)^2 + (1 - x_1^2)$.  For the optimizer, we use
simulated annealing~\cite{kirkpatrick1983optimization},
a gradient-free optimizer.  Simulated annealing will call the objective function
numerous times; for each simulation we limit the optimizer to 100K
objective evaluations.  We compare four frameworks%
%
\footnote{Another option here might be {\tt simulannealbnd()} 
in the Global Optimization Toolkit for MATLAB.
However, no license was available for these simulations.}
%
for this task:

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\tt ensmallen}
  \item {\tt scipy.optimize.anneal}, from scipy 0.14.1~\cite{jones2014scipy}
  \item simulated annealing implementation in {\tt Optim.jl} with Julia
1.0.1~\cite{mogensen2018optim}
  \item {\tt samin} in the {\tt optim} package for GNU Octave~\cite{octave}
\end{itemize}
\vspace*{-0.3em}

Table~\ref{tab:rosenbrock_results} shows the average runtime
across 10 trials for 100K iterations of simulated annealing with each
implementation.  The simulations were run on a % TODO: what are Marcus's
% system specs?
{\tt \$COMPUTER}.
Only Julia and {\tt ensmallen} are able to avoid the
function pointer dereference and take advantage of inlining and related
optimizations---and this difference is strongly reflected in the results.

%% CS: i'm not convinced that this conjecture is barking up the right tree;
%% CS: SciPy is in Python, which is by default an interpreted language.
%% CS: the slowdown many have nothing to do with inlining and simply due to
%% CS: Python being slow in general.
%% CS: it's possible to use PyPy or Numba to speed things up,
%% CS: in order to provide a "fairer" comparison against compiled languages.
%% CS: Octave also has an optional JIT compiler, which last time I checked
%% CS: needs to be explicitly enabled.
%% 
%% CS: i would suggest to stay away from the "we do inlining" argument,
%% CS: as it can be blown up too easily.
%% CS: furthermore, I would suggest to mention PyPy and/or Numba and Octave JIT
%% CS: (in order to show that we've done our homework),
%% CS: and then state something along the lines of
%% CS: "these approaches are not being used as we're comparing out-of-the-box performance"

Next, we consider the linear regression example first described in Section~\ref{sec:linreg_example}.
For this task we use the first-order L-BFGS optimizer~\cite{zhu1997algorithm},
so for {\tt ensmallen} we will implement two versions: one with {\tt Evaluate()}
and {\tt Gradient()}, and one with only {\tt EvaluateWithGradient()}.  The code
for each of these is the same as shown earlier.  In addition, we have more
options for Julia: we can implement only the objective function and allow either
{\tt Calculus.jl}\footnote{\url{https://github.com/JuliaMath/Calculus.jl}} or {\tt
ForwardDiff.jl}~\cite{RevelsLubinPapamarkou2016} packages to automatically
compute the gradient.  Overall, we compare against the following:

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\tt ensmallen}: only {\tt EvaluateWithGradient()} implemented
  \item {\tt ensmallen}-2: both {\tt Evaluate()} and {\tt Gradient()} separately
implemented
  \item {\tt Calculus.jl}: uses the default {\tt Calculus.jl} for
computing the gradient with {\tt Optim.jl}
  \item {\tt ForwardDiff.jl}: uses {\tt ForwardDiff.jl} for computing
the gradient with {\tt Optim.jl}
  \item {\tt Optim.jl}: implements the gradient manually
  \item {\tt scipy}: implements both objective and
gradient
  \item {\tt bfgsmin}: uses {\tt optim} package for GNU Octave
\end{itemize}
\vspace*{-0.3em}

Results for various data sizes are shown in Table~\ref{tab:lbfgs}.  For each
implementation, L-BFGS was allowed to run for only $10$ iterations and never
converged in fewer iterations.  The datasets used for training are highly noisy random
data with a slight linear pattern. Note that the exact data is not relevant
for the experiments here, only its size.  Runtimes are again reported as the
average across 10 runs.

\begin{table}
\begin{center}
\begin{tabular}{lccccc}
\toprule
{\em algorithm} & $d$: 100, $n$: 1k & $d$: 100, $n$: 10k & $d$: 100, $n$:
100k & $d$: 1k, $n$: 100k \\
\midrule
% TODO: this was only one trial on Ryan's desktop!
{\tt ensmallen} & {\bf 0.001s} & {\bf 0.006s} & {\bf 0.139s} & {\bf 1.698s} \\
{\tt ensmallen}-2 & 0.002s & 0.008s & 0.204s & 2.497s \\
{\tt Calculus.jl} & 0.172s & 0.960s & 27.428s & 2535.507s \\
{\tt ForwardDiff.jl} & 0.553s & 1.162s & 6.665s & 778.250s \\
{\tt Optim.jl} & 0.007s & 0.025s & 0.390s & 4.107s \\
{\tt scipy.optimize} & 0.002s & 0.010s & 0.198s & 2.022s \\
{\tt bfgsmin} & 0.215s & 1.705s & 15.998s & ??? \\
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for linear regression function with various dataset sizes,
with $n$ indicating the number of samples,
and $d$ indicating the dimensionality of each sample.
All Julia runs do not count compilation time.}
\label{tab:lbfgs}
\end{table}

The results indicate that {\tt ensmallen} with {\tt EvaluateWithGradient()}
is the fastest approach.
Furthremore, the use of {\tt EvaluateWithGradient()} yields considerable
speedup of almost 1.5x over the {\tt ensmallen} implementation with both the
objective and gradient implemented separately.  In addition, although the
automatic differentiation support makes it easier for the user to write their
code, it is clear that the code generated by the automatic differentiators
({\tt Calculus.jl} and {\tt ForwardDiff.jl})
is far from optimal, with runs that are slower by up to three orders of magnitude.
